Examining background gene lists in clusterProfiler - scRNA-seq example

### 2023-08-02 analysis/example_rnaseq.Rmd
write gene names out

@@

ora_dn$geneID <- NULL
ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
ora_dns <- rownames(ora_dn)

@@

orafix_up$geneID <- NULL
orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
orafix_ups <- rownames(orafix_up)

@@

orafix_dn$geneID <- NULL
orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
orafix_dns <- rownames(orafix_dn)

>>>>

writeLines(text=defup,con="aza_up.txt")
writeLines(text=defdn,con="aza_dn.txt")
writeLines(text=bg,con="aza_bg.txt")

@@

ora_dn$geneID <- NULL
ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=5)
ora_dns <- rownames(ora_dn)

@@

orafix_up$geneID <- NULL
orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=5)
orafix_ups <- rownames(orafix_up)

@@

orafix_dn$geneID <- NULL
orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=5)
orafix_dns <- rownames(orafix_dn)

### 2023-08-02 analysis/example_scrnaseq.Rmd
write gene names out

@@

def <- tbl[["CD4 T cells"]]

@@

defup <- subset(def,p_adj.loc<=0.05 & logFC>0)$gene

defdn <- subset(def,p_adj.loc<=0.05 & logFC<0)$gene

bg <- unique(def$gene)

message(paste("number of genes in background:",length(bg)))

@@

n_pw=10

ora_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_up$geneID <- NULL
ora_up <- subset(ora_up,p.adjust<0.05 & Count >=5)
ora_ups <- rownames(ora_up)

gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))

ora_up$es <- gr/br
ora_up <- ora_up[order(-ora_up$es),]
ora_up$Description=NULL

head(ora_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topup2 <- rev(head(ora_up$es,10))
names(topup2) <- rev(head(ora_up$ID,10))

@@

The reported background size does not match the dataset.

@@

ora_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_dn$geneID <- NULL
ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
ora_dns <- rownames(ora_dn)

gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))

ora_dn$es <- gr/br
ora_dn <- ora_dn[order(-ora_dn$es),]
ora_dn$Description=NULL

head(ora_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(ora_dn$es,n_pw)
names(topdn2) <- head(ora_dn$ID,n_pw)

Make a barplot

```{r,topbarplot1}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE KEGGs",
  xlab="ES",
  cex.main=0.9)

mtext("ORA test")

@@

```{r,subsetfix1}

defup <- unique(subset(def,p_adj.loc<=0.05 & logFC>0)$gene)

defdn <- unique(subset(def,p_adj.loc<=0.05 & logFC<0)$gene)

bg <- unique(def$gene)
message(paste("number of genes in background:",length(bg)))

@@

Adding all detected genes to the background appears to improve results!

```{r,gs2fix1}

bgdf <- data.frame("background",bg)
colnames(bgdf) <- c("term","gene")
genesets2 <- rbind(genesets2,bgdf)

```{r,oraupfix1}

n_pw=10

orafix_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

orafix_up$geneID <- NULL
orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
orafix_ups <- rownames(orafix_up)

gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))

orafix_up$es <- gr/br
orafix_up <- orafix_up[order(-orafix_up$es),]
orafix_up$Description=NULL

head(orafix_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topup2 <- rev(head(orafix_up$es,10))
names(topup2) <- rev(head(orafix_up$ID,10))

Now with the downregulated genes

```{r,oradnfix1}

orafix_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

orafix_dn$geneID <- NULL
orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
orafix_dns <- rownames(orafix_dn)

gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))

orafix_dn$es <- gr/br
orafix_dn <- orafix_dn[order(-orafix_dn$es),]
orafix_dn$Description=NULL

head(orafix_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(orafix_dn$es,n_pw)
names(topdn2) <- head(orafix_dn$ID,n_pw)

Make a barplot

```{r,topbarplot2fix1}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE KEGGs",
  xlab="ES",
  cex.main=0.9)

mtext("ORA fix test")

@@

v1 <- list("ORA up"=ora_ups, "ORA fix up"=orafix_ups)
v2 <- list("ORA dn"=ora_dns, "ORA fix dn"=orafix_dns)
v3 <- list("ORA"=union(ora_ups,ora_dns),
  "ORA fix"=union(orafix_ups,orafix_dns))

@@

ora <- c(ora_ups,ora_dns)
orafix <- c(orafix_ups,orafix_dns)

@@

```{r,examine1}

@@

ora_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_dn$geneID <- NULL
ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
ora_dns <- rownames(ora_dn)

gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))

ora_dn$es <- gr/br
ora_dn <- ora_dn[order(-ora_dn$es),]
ora_dn$Description=NULL

head(ora_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(ora_dn$es,n_pw)
names(topdn2) <- head(ora_dn$ID,n_pw)

Make a barplot

```{r,topbarplot22}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES",
  cex.main=0.9)

mtext("ORA test")

@@

Adding all detected genes to the background appears to improve results!

```{r,gs2fix2}

bgdf <- data.frame("background",bg)
colnames(bgdf) <- c("term","gene")
genesets2 <- rbind(genesets2,bgdf)

@@

```{r,oraupfix2}

n_pw=10

orafix_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

orafix_up$geneID <- NULL
orafix_up <- subset(orafix_up,p.adjust<0.05 & Count >=10)
orafix_ups <- rownames(orafix_up)

gr <- as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_up$BgRatio,"/"),"[[",2))

orafix_up$es <- gr/br
orafix_up <- orafix_up[order(-orafix_up$es),]
orafix_up$Description=NULL

head(orafix_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topup2 <- rev(head(orafix_up$es,10))
names(topup2) <- rev(head(orafix_up$ID,10))

Now with the downregulated genes

```{r,oradnfix2}

n_pw=10

orafix_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

orafix_dn$geneID <- NULL
orafix_dn <- subset(orafix_dn,p.adjust<0.05 & Count >=10)
orafix_dns <- rownames(orafix_dn)

gr <- as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(orafix_dn$BgRatio,"/"),"[[",2))

orafix_dn$es <- gr/br
orafix_dn <- orafix_dn[order(-orafix_dn$es),]
orafix_dn$Description=NULL

head(orafix_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in IFNB treatment") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(orafix_dn$es,n_pw)
names(topdn2) <- head(orafix_dn$ID,n_pw)

Make a barplot

```{r,topbarplot2fix2}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES",
  cex.main=0.9)

mtext("ORA fix test")

>>>>

```{r,orafunc}

ora <- function(fg,bg,genesets,n_pw=10,mincount=5) {

ora_res <- as.data.frame(enricher(gene = fg ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_res$geneID <- NULL
ora_res <- subset(ora_res,p.adjust<0.05 & Count >=mincount)
ora_res_sets <- rownames(ora_res)

gr <- as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",2))

ora_res$es <- gr/br
ora_res <- ora_res[order(-ora_res$es),]
ora_res$Description=NULL
return(ora_res)
}

orafix <- function(fg,bg,genesets,n_pw=10,mincount=5) {

bgdf <- data.frame("background",bg)
colnames(bgdf) <- c("term","gene")
genesets_fix <- rbind(genesets,bgdf)

ora_res <- as.data.frame(enricher(gene = fg ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets_fix,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora_res$geneID <- NULL
ora_res <- subset(ora_res,p.adjust<0.05 & Count >=mincount)
ora_res_sets <- rownames(ora_res)

gr <- as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_res$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora_res$BgRatio,"/"),"[[",2))

ora_res$es <- gr/br
ora_res <- ora_res[order(-ora_res$es),]
ora_res$Description=NULL
return(ora_res)
}

@@

@@

def <- lapply(tbl , function(x) {
  bg <- unique(x$gene)
  up <- unique(subset(x,p_adj.loc<0.05&logFC>0)$gene)
  dn <- unique(subset(x,p_adj.loc<0.05&logFC<0)$gene)
  return(list("bg"=bg,"up"=up,"dn"=dn))
} )

str(def)

@@

ups <- lapply(1:length(def),function(i) {
  ora(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
})

up_sets <- unlist(lapply(1:length(ups) , function(i) {
  nm <- names(def)[[i]]
  sets <- ups[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(up_sets)
up_sets

@@

@@

dns <- lapply(1:length(def),function(i) {
  ora(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
})

dn_sets <- unlist(lapply(1:length(dns) , function(i) {
  nm <- names(def)[[i]]
  sets <- dns[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(dn_sets)
dn_sets

@@

```{r,orafixup1}

fixups <- lapply(1:length(def),function(i) {
  orafix(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
})

fixup_sets <- unlist(lapply(1:length(fixups) , function(i) {
  nm <- names(def)[[i]]
  sets <- fixups[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(fixup_sets)
fixup_sets

@@

```{r,orafixdn1}

fixdns <- lapply(1:length(def),function(i) {
  orafix(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
})

fixdn_sets <- unlist(lapply(1:length(fixdns) , function(i) {
  nm <- names(def)[[i]]
  sets <- fixdns[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(fixdn_sets)
fixdn_sets

@@

v1 <- list("ORA up"=up_sets, "ORA fix up"=fixup_sets)
v2 <- list("ORA dn"=dn_sets, "ORA fix dn"=fixdn_sets)
v3 <- list("ORA"=union(up_sets,dn_sets),
  "ORA fix"=union(fixup_sets,fixdn_sets))

@@

ora <- c(up_sets,dn_sets)
orafix <- c(fixup_sets,fixdn_sets)

@@

@@

dns <- lapply(1:length(def),function(i) {
  ora(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
})

dn_sets <- unlist(lapply(1:length(dns) , function(i) {
  nm <- names(def)[[i]]
  sets <- dns[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(dn_sets)
dn_sets

@@

```{r,orafixup2}

fixups <- lapply(1:length(def),function(i) {
  orafix(def[[i]]$up,def[[i]]$bg,genesets=genesets2)
})

fixup_sets <- unlist(lapply(1:length(fixups) , function(i) {
  nm <- names(def)[[i]]
  sets <- fixups[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(fixup_sets)
fixup_sets

@@

```{r,orafixdn2}

fixdns <- lapply(1:length(def),function(i) {
  orafix(def[[i]]$dn,def[[i]]$bg,genesets=genesets2)
})

fixdn_sets <- unlist(lapply(1:length(fixdns) , function(i) {
  nm <- names(def)[[i]]
  sets <- fixdns[[i]]$ID
  if ( length(sets) >0 ) { paste(nm,sets) }
} ) )

length(fixdn_sets)
fixdn_sets
